---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

## Loading and preprocessing the data
First unzip the data to the working directory. Then read the data to a dataframe using read.csv function.
```{r loading data, eval=TRUE}
setwd("/Users/Brandon/datasciencecoursera/")
unzip("./RepData_PeerAssessment1/activity.zip")
data <- read.csv("./activity.csv",header=T,sep=",")
options(scipen = 10, digits = 0)
```

## What is mean total number of steps taken per day?
Use tapply to calculate the sum of number of steps taken each day. Then calculate the mean and median. Create a histogram with the Date on the x-axis and the value of daily sum on the y-axis.
```{r First question}
dailysum<-tapply(data$steps,data$date,sum,na.rm=T)
dailymean<-mean(dailysum)
dailymedian<-median(dailysum)
dailymeandate<-as.Date(names(dailysum))
library(ggplot2)
qplot(dailymeandate,dailysum,geom="histogram",stat="identity")
```

The mean total number of steps taken per day is `r round(dailymean,0)` and the median is `r round(dailymedian,0)`

## What is the average daily activity pattern?
Use tapply to calculate the mean of number of steps taken at each interval across all days.
```{r Second question}
interval_mean<-tapply(data$steps,data$interval,mean,na.rm=T)
time_in_a_day<-names(interval_mean)
maxint<-names(interval_mean[interval_mean==max(interval_mean)])
plot(time_in_a_day,interval_mean,type="l")
```

The maximum umber of steps taken by the individual starts at `r maxint` during the day


## Imputing missing values
Fill in the NA values with the mean conditional of the 5-minute interval. For instance, a NA value occurs at 0000 hours will be filled with the mean of all the non-NA values of the other days at 0000 hours
```{r Third question part 1}
NoNAs<-length(data$steps[data$steps=="NA"])
data2<-data
for (i in 1:length(data2$steps)){
        if (is.na(data2$steps[i])){
                data2$steps[i]<-interval_mean[names(interval_mean)==data2$interval[i]]
        }}
```
There are `r NoNAs` NA's in the "steps" data.

```{r Third question part 2}
dailysum2<-tapply(data2$steps,data2$date,sum)
dailymean2<-mean(dailysum2)
dailymedian2<-median(dailysum2)
dailymeandate2<-as.Date(names(dailysum2))
qplot(dailymeandate2,dailysum2,geom="histogram",stat="identity")

```

After filling in the missing data, the mean total number of steps taken per day is now `r round(dailymean2,0)` and the median is `r round(dailymedian2,0)`. These are different from the original mean, `r dailymean`, and median, `r dailymedian`.

## Are there differences in activity patterns between weekdays and weekends?

Convert the date to a weekday variable and then categorize them to either "weekday" or "weekend" in a new variable weekdays2.
```{r Fourth question part 1}
data2$weekdays<-weekdays(as.Date(data2$date),abbreviate = TRUE)
data2$weekdays2<-"weekday"
data2$weekdays2[data2$weekdays=="Sat"|data2$weekdays=="Sun"]<-"weekend"
```

Create two subsets of data, one for weekday and one for weekend, and then create the panel chart.
```{r Fourth question part 2}
weekday_data2<-subset(data2,data2$weekdays2=="weekday")
weekend_data2<-subset(data2,data2$weekdays2=="weekend")
weekday_interval_mean<-tapply(weekday_data2$steps,weekday_data2$interval,mean)
weekend_interval_mean<-tapply(weekend_data2$steps,weekend_data2$interval,mean)

par(mfrow=c(2,1),mar=c(4,4,2,1))
plot(time_in_a_day,weekend_interval_mean,type="l")
plot(time_in_a_day,weekday_interval_mean,type="l")
```

The sample data shows a lot more activities during weekend compared to that of weekday, especially after 10am in the day.

